---
title: "microRNA Trophoblast Analysis"
author: "Magda"
date: "27 11 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
require(knitr)
opts_knit$set(root.dir="C:/Users/x1/Desktop/part2_analysis/microrna")
```

## Set of Analysis

Change working directory to where the data for the mRNA analysis is situated.
Load required packages for the analysis.

```{r}
setwd("C:/Users/x1/Desktop/part2_analysis/microrna")
library(DESeq2)
library(ggplot2)
library(Rtsne)
library(RColorBrewer)
library(gplots)

```

## Including Plots

Read in the sample description file *pdata.txt* as an object called pdata.

```{r pressure, echo=FALSE}
pdata = read.table("pdata.txt",header=TRUE,sep="\t")
pdata$contrast=paste(pdata$culture,"_",pdata$class,sep="")
pdata
c19mc=read.table("c19mc_mirs.txt",header=F)
c19mc=as.character(c19mc$V1)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


```{r}
conds = as.factor(pdata$sampletype)

condcols=brewer.pal(n = length(levels(conds)), name = 'Dark2')
names(condcols)=levels(conds)

classcols = brewer.pal(n = length(levels(as.factor(pdata$classtype))), name = 'Set3')
names(classcols)=levels(as.factor(pdata$classtype))

culturecols = c("red","darblue")
names(culturecols)=c("2D","3D")
```

## Data Loading and Normalisation via DESeq2

```{r}

mircounts=read.table("microrna.counts",header=T,row.names=1)
reorder=c(8,5,7,4,1,3,12,9,11,16,13,15,20,17,19,6,2,10,14,18,25,24,27,26,29,28,22,21,30,31,33,34,32,23)

mircounts=mircounts[,reorder]
mircounts = mircounts[,c(1:15,21:28)]
colnames(mircounts)=pdata$sampleno

ddsHTSeq = DESeqDataSetFromMatrix(as.matrix(mircounts),colData = pdata,design =~ contrast)

# Data Normalisation
dds = estimateSizeFactors(ddsHTSeq)
# Data Dispersion
dds = estimateDispersions(dds)
# New Neg. Binomial test
dds = nbinomWaldTest(dds)



counts_table_raw=counts(dds,normalized=FALSE)
counts_table_norm=counts(dds,normalized=TRUE)
plotDispEsts(dds,main='Dispersion Plot')

```

# Sample QC
```{r}
length_data=read.table("lengths.txt",header=T,row.names=1)

 


names=colnames(length_data)
names=gsub("_R1_merged.fastq.gz.tally.fasta","",names)

 

cols=rainbow(length(names))
names(cols)=names

 

matplot(length_data,type="l",lty=1,col=cols,xlim=c(0,70),ylab="No Reads",xlab="Length")
legend("topright",fill = cols,legend = names,cex=0.4)

 


for (i in 1:length(names)){
  plot(length_data[,i],type="l",col=cols[i],main=names[i],ylab="No Reads",xlab="Length")
}

## Number of reads per sample


barplot(apply(counts_table_raw,2,sum),las=2,col=condcols[pdata$sampletype],main="Raw Counts",cex.names = 0.6)
legend("bottomleft",levels(as.factor(pdata$sampletype)),fill=condcols[levels(as.factor(pdata$sampletype))],cex=0.4)
barplot(apply(counts_table_norm,2,sum),las=2,col=condcols[pdata$sampletype],main="Norm Counts",cex.names=0.6)
legend("bottomleft",levels(as.factor(pdata$sampletype)),fill=condcols[levels(as.factor(pdata$sampletype))],cex=0.4)


```

## Correlation Analysis
```{r}
hmcol = colorRampPalette(brewer.pal(9, 'GnBu'))(100)
heatmap.2(cor(counts_table_norm),trace="none",col=hmcol,labRow = pdata$sampleno,cexRow=0.5,RowSideColors = condcols[pdata$sampletype],ColSideColors = classcols[pdata$classtype],main="Normalised Heatmap",labCol=pdata$classtype,cexCol = 0.5)
```

## Perform VST transformation to stabilise low counts


```{r}
vsd = varianceStabilizingTransformation(dds)
vstMat = assay(vsd)
heatmap.2(cor(vstMat),trace="none",col=hmcol,labRow = pdata$sampleno,cexRow=0.5,RowSideColors = condcols[pdata$sampletype],ColSideColors = classcols[pdata$classtype],main="Normalised Heatmap",labCol=pdata$classtype,cexCol = 0.5)
```

## Multidimensional Projectsion (PCA, tSNE)

```{r}
pca = princomp(assay(vsd))
plot(pca$loadings, main='PCA Variance Stabilised', col=condcols[pdata$sampletype],  pch=19, cex=1)
text(pca$loadings, as.character(conds), pos=3, cex=0.8)

pca = princomp(assay(vsd))
plot(pca$loadings, main='PCA Variance Stabilised', col=classcols[pdata$classtype],  pch=19, cex=1)
text(pca$loadings, as.character(pdata$classtype), pos=3, cex=0.8)
```
```{r}
set.seed(46)
tsne = Rtsne(t(vstMat),perplexity=5,check_duplicates = FALSE)

tsne.df <- data.frame(tsne.1 = tsne$Y[,1], tsne.2 = tsne$Y[,2])
ggplot(data = tsne.df, aes(tsne.1, tsne.2)) +
  geom_point(size = 4, pch = 20, colour = condcols[conds]) +
  geom_text(size = 1.5, vjust=2, aes(label=pdata$sampleno)) +
  scale_alpha_discrete(range = c(0.5, 1)) +
  theme_minimal() +
  ylab("tSNE 1") +
  xlab("tSNE 2")

tsne.df <- data.frame(tsne.1 = tsne$Y[,1], tsne.2 = tsne$Y[,2])
ggplot(data = tsne.df, aes(tsne.1, tsne.2)) +
  geom_point(size = 4, pch = 20, colour = classcols[pdata$classtype]) +
  geom_text(size = 1.5, vjust=2, aes(label=pdata$sampleno)) +
  scale_alpha_discrete(range = c(0.5, 1)) +
  theme_minimal() +
  ylab("tSNE 1") +
  xlab("tSNE 2")  
```

# Statistical Analysis

Set up the contrasts desired
```{r}

# Set statistical thresholds
lfc_threshold = 1.2
p_value_threshold = 0.05
c19mc_common=intersect(rownames(counts(dds)),c19mc)
 


# Perform the 2D comparison
res2D <- results(dds, contrast=c("contrast","2D_villous","2D_extravillous"))
res2D$genename=rownames(res2D)
res2D=res2D[order(res2D$padj),]
write.table(res2D,file = "res2dmi.txt",sep="\t",quote=F)
sig_hits_2D = rownames(res2D[(res2D$padj<= p_value_threshold) & (!is.na(res2D$padj)) & (abs(res2D$log2FoldChange)>= lfc_threshold) ,])

plot(res2D$log2FoldChange,-log10(res2D$padj),pch=19,cex=0.3,col="darkblue",main="Volcano Plot\n2D villous v extravillous")
points(res2D[sig_hits_2D,]$log2FoldChange,-log10(res2D[sig_hits_2D,]$padj),pch=19,cex=0.3,col="red")
points(res2D[c19mc_common,]$log2FoldChange,-log10(res2D[c19mc_common,]$padj),pch=20,cex=0.3,col="green")
text(res2D[sig_hits_2D,]$log2FoldChange[1:30],-log10(res2D[sig_hits_2D,]$padj[1:30]),labels =sig_hits_2D,pos=2,cex=0.4)
abline(h=-log10(p_value_threshold),lty=2,col="red")
abline(v=c(-lfc_threshold,lfc_threshold),lty=2,col="red")

# Perform the 3D comparison
res3D <- results(dds, contrast=c("contrast","3D_villous","3D_extravillous"))
res3D$genename=rownames(res3D)
res3D=res3D[order(res3D$padj),]
write.table(res3D,file = "res3dmi.txt",sep="\t",quote=F)
sig_hits_3D = rownames(res3D[(res3D$padj<= p_value_threshold) & (!is.na(res3D$padj)) & (abs(res3D$log2FoldChange)>= lfc_threshold) ,])

plot(res3D$log2FoldChange,-log10(res3D$padj),pch=19,cex=0.3,col="darkblue",main="Volcano Plot\n3D villous v extravillous")
points(res3D[sig_hits_3D,]$log2FoldChange,-log10(res3D[sig_hits_3D,]$padj),pch=19,cex=0.3,col="red")
points(res3D[c19mc_common,]$log2FoldChange,-log10(res3D[c19mc_common,]$padj),pch=20,cex=0.3,col="green")
text(res3D[sig_hits_3D,]$log2FoldChange[1:30],-log10(res3D[sig_hits_3D,]$padj[1:30]),labels = sig_hits_3D,pos=2,cex=0.4)
abline(h=-log10(p_value_threshold),lty=2,col="red")
abline(v=c(-lfc_threshold,lfc_threshold),lty=2,col="red")

```